---
title: "Epitope Prediction Project"
author: 
    - "Priyanshu Dey"
    - "Adam Torres Encarnacion"
    - "Elizabeth Marchini"
date: "`r Sys.Date()`"
output: html_document
---

# Epitope Prediction Using Generalized Linear Models and Multilevel Models

This report outlines the analysis and modeling of the epitope prediction dataset, structured into two main parts: Generalized Linear Models (GLMs) and Multilevel Modeling.

## Table of Contents
1. [Introduction](#introduction)
2. [Exploratory Data Analysis (EDA)](#eda)
3. [Part 1: Generalized Linear Models (GLMs)](#glm)
4. [Part 2: Multilevel Modeling](#multilevel)
5. [Conclusion](#conclusion)
6. [Appendix](#appendix)

## Introduction {#introduction}


## Exploratory Data Analysis (EDA) {#eda}
###

```{r}
# clear workspace
rm(list = ls()) 
# load libraries
library(tidyverse)

```
### Data Importing 
Bcells colnames
```{r}
bcells <- read.csv("Data/input_bcell.csv")
colnames(bcells)
```
Covid colnames
```{r}
covid <- read.csv("Data/input_covid.csv")
colnames(covid)
```
Sars colnames
```{r}
sars <- read.csv("Data/input_sars.csv")
colnames(sars)
```
### Data Cleaning

### Data Description
Our chosen dataset – COVID-19/SARS B-cell Epitope Prediction – analyzes antibody-inducing activity given 10 features/predictors. This dataset was developed as part of Future’s research endeavors in 2020 and served as a way for data analysts to get involved in the search for a cure for the COVID-19 virus. It has since been made available on Kaggle. The dataset describes whether an amino acid peptide exhibited antibody-inducing activity (marked by an activity label). The original dataset contained five categories, including four classes of “Positive,” which the researchers combined into one. This attributes the task to a simple binary classification, suitable for this project.


Chosen dataset: [COVID-19/SARS B-cell Epitope Prediction](https://www.kaggle.com/datasets/futurecorporation/epitope-prediction/data)


Variables
• start_position (peptide): quantitative value of the number of peptides into the sequence the protein starts
• end_position (peptide): quantitative value of the number of peptides into the sequence the protein ends
• chou_fasman (peptide): quantitative value of the beta turns feature from peptide information
• emini (peptide): quantitative value of the surface accessibility feature from peptide information
• kolaskar_tongaonkar (peptide): quantitative value of the antigenicity feature from peptide information
• parker (peptide): quantitative value of the hydrophobicity feature from peptide information
• isoelectric_point (protein): quantitative value of the isoelectric point of the protein 
• aromacity (protein): quantitative value of the aromaticity of the protein 
• stability (protein): quantitative value of the stability of the protein

## Part 1: Generalized Linear Models (GLMs) {#glm}

### Research Question and Hypotheses
How accurately can amino acid peptides and parent proteins predict whether a peptide exhibits antibody-inducing activity (antibody valence = 1) in the context of COVID- 19/SARS vaccine development? 


### Exploratory Data Analysis

```{r}
#things Elizabeth did to just share with the group
#first thing I did was rename the some of the variables for easier typing! if you guys dont
#want this no worries
bcells <-
  bcells %>%
  rename(kt = kolaskar_tongaonkar, 
         start = start_position,
         end = end_position,
         iso = isoelectric_point,
         aroma = aromaticity,
         hydro = hydrophobicity) %>%
  mutate(pep_length = (end - start))

sars <- sars %>%
  rename(kt = kolaskar_tongaonkar, 
         start = start_position,
         end = end_position,
         iso = isoelectric_point,
         aroma = aromaticity,
         hydro = hydrophobicity) %>%
  mutate(pep_length = (end - start))


covid <- covid %>%
  rename(kt = kolaskar_tongaonkar, 
         start = start_position,
         end = end_position,
         iso = isoelectric_point,
         aroma = aromaticity,
         hydro = hydrophobicity) %>%
  mutate(pep_length = (end - start))


#i made a basic linear model and did some EDA Plots
model1 <- lm(target ~ chou_fasman + emini + kt + parker + iso + aroma + hydro + stability + pep_length, data = bcells)
summary(model1)

plot(model1, which = c(1:2,4))



confint(model1, level = 0.95)



```
Residual vs Fitted analysis: 
The residuals appear to have a funnel shape, suggesting hetroscdasticy where the variability of the Residual is not constant across the fitted values. This is a violation of the assumption of homoscedasticity.

Q-Q Plot analysis: 
The plot shows that there is a significant deviation from the diagonal line, suggesting that the residuals are not normally distributed. This is a violation of the assumption of normality.

Cook's Distance: 
The cooks distance shows that there are 3 influential points in the data such as 8469, 9892, and 11326. These points are likely to have a large impact on the model.


### Model Analysis


### Results


### Discussion


## Part 2: Multilevel Modeling {#multilevel}

The data set chosen for part 2 is the reaction time dataset, which observes the reaction time for people in the study to react to a change in a picture viewing task.


Variables

Response: Reaction time
Level 1 Observational Unit: Picture they were observing
Level 2 Observational Unit:  Each person in the study
Level 1 Variables: meaning, salience, c_mean, c_sal
Level 2 Variables: id, sex, age,oldage, yrs65
Fixed Effect: old age, salience
Random Effect: Person in study (id)



### Research Question and Hypotheses

What is the association with painting salience and reaction time?

Does the association between painting salience and reaction time differ based on old age?



### Data Structuring 

```{r}
read.csv("https://raw.githubusercontent.com/mkshaw/learn-mlms/master/data/hoffman2007.csv")

```



### Model Building


### Results



### Discussion


## Conclusion {#conclusion}



## Appendix {#appendix}




### References


