---
title: "Epitope Prediction Project"
author: 
    - "Priyanshu Dey"
    - "Adam Torres Encarnacion"
    - "Elizabeth Marchini"
date: "`r Sys.Date()`"
output: html_document
---

# Epitope Prediction Using Generalized Linear Models and Multilevel Models

## Introduction {#introduction}
This report outlines the analysis and modeling of the epitope prediction dataset, structured into two main parts: Generalized Linear Models (GLMs) and Multilevel Modeling.

## Table of Contents
1. [Introduction](#introduction)
2. [Part 1: Generalized Linear Models (GLMs)](#glm)
3. [Part 2: Multilevel Modeling](#multilevel)
4. [Conclusion](#conclusion)
5. [Appendix](#appendix)


### Research Question and Hypotheses

How accurately can amino acid peptides and parent proteins predict whether a peptide exhibits antibody-inducing activity (antibody valence = 1) in the context of COVID-19/SARS vaccine development?

**Hypotheses:** 

- **Null Hypothesis (H₀):** Amino acid peptides and parent proteins do not significantly predict antibody-inducing activity.
- **Alternative Hypothesis (H₁):** Amino acid peptides and parent proteins significantly predict antibody-inducing activity.

## Exploratory Data Analysis (EDA) {#eda}
### Library Loading and Data Import

```{r}
# clear workspace
rm(list = ls()) 
# load libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(MASS)
library(GGally)
library(gridExtra)
library(lme4)

```

### Data Importing and Cleaning
```{r}
bcells <- read.csv("Data/input_bcell.csv")

# Rename columns and calculate peptide length
bcells <- bcells %>%
  rename(kt = kolaskar_tongaonkar, 
         start = start_position,
         end = end_position,
         iso = isoelectric_point,
         aroma = aromaticity,
         hydro = hydrophobicity) %>%
  mutate(pep_length = (end - start))

# Ensure target is valid for Poisson
bcells <- bcells %>% mutate(target = pmax(target, 0))

# Summary statistics
summary(bcells)
```




## Part 1: Generalized Linear Models (GLMs) {#glm}


# Residual Diagnostics

### Poisson GLM Modeling

We initially fit a Poisson GLM:

\[
\text{log}(E[\text{target}]) = \beta_0 + \beta_1(\text{chou_fasman}) + \beta_2(\text{emini}) + \dots + \beta_9(\text{pep_length})
\]
#### Poisson Model Fitting {#poisson-glm-modeling}
```{r}
# Poisson regression model
poisson_model <- glm(target ~ chou_fasman + emini + kt + parker + iso + aroma + hydro + stability + pep_length, 
                     family = poisson(link = "log"), data = bcells)

# Model summary
summary(poisson_model)

# Check for Overdispersion
pearson_residuals <- residuals(poisson_model, type = "pearson")
overdispersion <- sum(pearson_residuals^2) / poisson_model$df.residual
cat("Overdispersion Statistic:", round(overdispersion, 2), "\n")
```

#### : Residual Diagnostics and QQ Plots {#residual-diagnostics-and-qq-plots}
```{r}
# 1. Residuals vs Fitted
plot(poisson_model$fitted.values, residuals(poisson_model, type = "pearson"),
     xlab = "Fitted Values", ylab = "Pearson Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# 2. QQ Plot of Residuals
qqnorm(residuals(poisson_model, type = "pearson"), main = "QQ Plot of Residuals")
qqline(residuals(poisson_model, type = "pearson"), col = "red")

# 3. Scale-Location Plot
plot(poisson_model$fitted.values, sqrt(abs(residuals(poisson_model, type = "pearson"))),
     xlab = "Fitted Values", ylab = "Sqrt(|Pearson Residuals|)",
     main = "Scale-Location Plot")
abline(h = 0, col = "red", lty = 2)

# 4. Histogram of Residuals
# Residuals vs Fitted Plot
plot(poisson_model$fitted.values, residuals(poisson_model, type = "pearson"),
  main = "Residuals vs Fitted Plot for Poisson Model",
  xlab = "Fitted Values",
  ylab = "Pearson Residuals",
  pch = 20, col = "blue")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))  # Reset layout
```


#### Results
**Residuals vs Fitted Plot**
The residuals show a curved pattern, suggesting a possible violation of the mean-variance assumption of the Poisson model.
The systematic structure indicates that the model might not fully capture the relationship between predictors and the target variable.
**QQ Plot of Residuals**
Residuals deviate significantly from the red line, particularly in the tails.
This suggests that the residuals are not normally distributed, which is expected in a Poisson model, as it assumes count data, not normality.
**Scale-Location Plot**
The pattern in the plot suggests non-constant variance, as residuals are unevenly spread across fitted values.
This confirms that the variance increases with the fitted values, a common issue with Poisson models.
**Histogram of Pearson Residuals**
The residuals are centered around zero but exhibit skewness, particularly with more extreme positive values.
This indicates that some observations may have higher-than-expected counts, contributing to the model’s limitations.
#### Conclusion
The Poisson GLM assumptions are not fully met, as evidenced by the residual diagnostics. The model shows signs of overdispersion, non-constant variance, and non-normality in the residuals. Further model refinement or alternative modeling approaches may be necessary to improve the fit and predictive performance.

### Negative Binomial GLM {#negative-binomial-glm}
To address overdispersion, we fit a Negative Binomial model:

\[
\text{log}(E[\text{target}]) = \beta_0 + \beta_1(\text{chou_fasman}) + \dots + \beta_9(\text{pep_length})
\]

#### Negative Binomial Model Fitting 
```{r}
# Load required library for Negative Binomial

# Fit Negative Binomial model
nb_model <- glm.nb(target ~ chou_fasman + emini + kt + parker + iso + aroma + hydro + stability + pep_length, 
                   data = bcells)

# Model summary
summary(nb_model)

# Extract and exponentiate coefficients to get rate ratios
nb_coef <- exp(coef(nb_model))
nb_results <- data.frame(Predictor = names(nb_coef),
                         Exp_Coefficient = nb_coef,
                         Coefficient = coef(nb_model))
print(nb_results)


```

#### Residual Diagnostics and QQ Plots {#residual-diagnostics-and-qq-plots-nb}
```{r}
# Set up a 2x2 plot layout for diagnostics
par( mfrow= c(2, 2))

# 1. Residuals vs Fitted
plot(nb_model$fitted.values, residuals(nb_model, type = "pearson"),
     xlab = "Fitted Values", ylab = "Pearson Residuals",
     main = "Residuals vs Fitted", pch = 20, col = "black")
abline(h = 0, col = "red", lty = 2)

# 2. QQ Plot of Residuals
qqnorm(residuals(nb_model, type = "pearson"), main = "QQ Plot of Residuals", pch = 20)
qqline(residuals(nb_model, type = "pearson"), col = "red", lwd = 2)

# 3. Scale-Location Plot
plot(nb_model$fitted.values, sqrt(abs(residuals(nb_model, type = "pearson"))),
     xlab = "Fitted Values", ylab = "Sqrt(|Pearson Residuals|)",
     main = "Scale-Location Plot", pch = 20, col = "black")
abline(h = 0, col = "red", lty = 2)

# 4. Residuals vs Fitted Plot for Negative Binomial Model
plot(nb_model$fitted.values, residuals(nb_model, type = "pearson"),
     xlab = "Fitted Values", 
     ylab = "Pearson Residuals", 
     main = "Residuals vs Fitted for Negative Binomial Model",
     pch = 20, col = "black")

```

### Results

**Residuals vs Fitted Plot**
The residuals show a curved pattern, suggesting a possible violation of the mean-variance assumption of the Negative Binomial model.
The systematic structure indicates that the model might not fully capture the relationship between predictors and the target variable.
**QQ Plot of Residuals**
Residuals deviate significantly from the red line, particularly in the tails.
This suggests that the residuals are not normally distributed, which is expected in a Negative Binomial model, as it assumes count data, not normality.
**Scale-Location Plot**
The pattern in the plot suggests non-constant variance, as residuals are unevenly spread across fitted values.
This confirms that the variance increases with the fitted values, a common issue with Negative Binomial models.
**Histogram of Pearson Residuals**
The residuals are centered around zero but exhibit skewness, particularly with more extreme positive values.
This indicates that some observations may have higher-than-expected counts, contributing to the model’s limitations.


Initially, the Poisson GLM revealed patterns in the residual diagnostics—such as deviations in QQ plots and residuals vs. fitted plots—that suggested potential overdispersion and violations of key model assumptions. These issues indicated that a standard Poisson model may not fully capture the data’s underlying structure.

Transitioning to a Negative Binomial model, which is more flexible in handling overdispersed count data, improved some aspects of model fit but still did not completely resolve the distributional challenges. Nevertheless, across both modeling attempts, certain predictors consistently emerged as significant. Notably, the chou_fasman and hydro predictors showed strong positive associations with the probability of inducing an antibody response. Additional variables, such as kt, iso, and pep_length, also demonstrated statistically significant but more moderate effects.

From a hypothesis-testing perspective, these results lead us to reject the null hypothesis for key predictors, confirming that certain physicochemical and structural properties of peptides do influence immunogenicity. However, the persistent complexity in the residual diagnostics suggests that traditional GLMs—even with Negative Binomial adjustments—may only partially capture the true underlying relationships.

In conclusion, this analysis supports the importance of specific peptide-level features in predicting antibody-inducing activity.` significantly increase the likelihood of antibody-inducing activity, making them critical predictors in epitope prediction.

## Part 2: Multilevel Modeling {#multilevel}

```{r}
rtdata <- read.csv("https://raw.githubusercontent.com/mkshaw/learn-mlms/master/data/hoffman2007.csv")

#need to make salience a 01 type data structure; between 1 and 5 so maybe if painting salience is less than 2.5 and greater than equal to 2.5
rtdata <- rtdata %>%
          mutate(salience = ifelse(salience >= 2.5, 1,0))
```

```{r}
# Select numeric variables for the scatterplot matrix
numeric_vars <- rtdata %>%
                dplyr::select(age, salience, rt_sec, meaning, c_mean, c_sal)

# Create the scatterplot matrix
ggpairs(numeric_vars, 
        aes(color = as.factor(salience), alpha = 0.7)) +
        theme_minimal()
```

```{r}
#random intercepts model
RImodel <- lmer(rt_sec ~ 1 + (1|id), data = rtdata)

summary(RImodel)
```
```{r}
ggplot(data = data.frame(residuals = resid(RImodel)), aes(sample = residuals)) +
stat_qq() +
stat_qq_line() +
theme_minimal() +
labs(title = "QQ Plot of Residuals",
     x = "Theoretical Quantiles",
     y = "Sample Quantiles")
```

Inter-class correlation coefficient calculation
```{r}
var_hat <- 53.44
var_uhat <- 12.31
  
phat <- var_uhat/ (var_uhat+var_hat)
# the closer this number is to 1, the more evidence we have to use a multilevel model
print(phat)
```

Our fixed effect parameter was 7.4402, which represents the average reaction time across all individuals all ages.

Our inter-class correlation Coefficient Calculation showed that 18.72% of total variation in reaction time score is due to the difference in person.

Random Slopes and Intercept model

```{r}
RSImodel <- lmer(rt_sec ~ salience + (1 + salience|id), REML = TRUE, data = rtdata)
summary(RSImodel)
```

Based on the random slopes and intercepts model, we expect reaction time to decrease by 1.09984 on average for images with a salience greater than 2.5. The more salient a picture is means the more obvious the change will be so this indicates that as the image get clearer the reaction time decreases. 

```{r}
ggplot(data = data.frame(Fitted = fitted(RSImodel), Residuals = resid(RSImodel)), 
                         aes(x = Fitted, y = Residuals)) +
  
geom_point(alpha = 0.6, color = "#112446") +
geom_hline(yintercept = 0, color = "red") +
theme_minimal() +
labs(title = "Residuals vs. Fitted Values",
     x = "Fitted Values",
     y = "Residuals")
```

Model 3- REML = False
```{r}
RSImodel <- lmer(rt_sec ~ salience + oldage + oldage:salience + (1 + salience|id), REML = FALSE, data = rtdata)
summary(RSImodel)

AIC(RSImodel)
BIC(RSImodel)
```

```{r}
ggplot(data = data.frame(residuals = resid(RSImodel)), aes(sample = residuals)) +
stat_qq() +
stat_qq_line() +
theme_minimal() +
labs(title = "QQ Plot of Residuals",
     x = "Theoretical Quantiles",
     y = "Sample Quantiles")
```