---
title: "Epitope Prediction Project"
author: 
    - "Priyanshu Dey"
    - "Adam Torres Encarnacion"
    - "Elizabeth Marchini"
date: "`r Sys.Date()`"
output: html_document
---

# Epitope Prediction Using Generalized Linear Models and Multilevel Models

This report outlines the analysis and modeling of the epitope prediction dataset, structured into two main parts: Generalized Linear Models (GLMs) and Multilevel Modeling.

## Table of Contents
1. [Introduction](#introduction)
2. [Exploratory Data Analysis (EDA)](#eda)
3. [Part 1: Generalized Linear Models (GLMs)](#glm)
4. [Part 2: Multilevel Modeling](#multilevel)
5. [Conclusion](#conclusion)
6. [Appendix](#appendix)

## Introduction {#introduction}


## Exploratory Data Analysis (EDA) {#eda}
###

```{r}
# clear workspace
rm(list = ls()) 
# load libraries
library(tidyverse)

```
### Data Importing 
```{r}
cat("COVID-19 B-Cell Epitope Prediction Data")
cat("bcells data")
bcells <- read.csv("Data/input_bcell.csv")
colnames(bcells)
cat("covid data")
covid <- read.csv("Data/input_covid.csv")
colnames(covid)
cat("sars data")
sars <- read.csv("Data/input_sars.csv")
colnames(sars)
```
### Data Cleaning

### Data Description
Our chosen dataset – COVID-19/SARS B-cell Epitope Prediction – analyzes antibody-inducing activity given 10 features/predictors. This dataset was developed as part of Future’s research endeavors in 2020 and served as a way for data analysts to get involved in the search for a cure for the COVID-19 virus. It has since been made available on Kaggle. The dataset describes whether an amino acid peptide exhibited antibody-inducing activity (marked by an activity label). The original dataset contained five categories, including four classes of “Positive,” which the researchers combined into one. This attributes the task to a simple binary classification, suitable for this project.
Chosen dataset: [COVID-19/SARS B-cell Epitope Prediction](https://www.kaggle.com/datasets/futurecorporation/epitope-prediction/data)
Dataset analysis-  due december 3rd
Variables
start_position (peptide)
• end_position (peptide)
• chou_fasman (peptide)
• emini (peptide)
• kolaskar_tongaonkar (peptide)
• parker (peptide)
• isoelectric_point (protein)
• aromacity (protein)
• stability (protein)

## Part 1: Generalized Linear Models (GLMs) {#glm}

### Research Question and Hypotheses
How accurately can amino acid peptides and parent proteins predict whether
a peptide exhibits antibody-inducing activity (antibody valence = 1) in the context of COVID- 19/SARS vaccine development? 


### Model Building



### Results


### Discussion


## Part 2: Multilevel Modeling {#multilevel}

### Research Question and Hypotheses


### Data Structuring


### Model Building


### Results



### Discussion


## Conclusion {#conclusion}



## Appendix {#appendix}




### References


