## PART 2

```{r}
library(tidyverse)
library(GGally)
library(gridExtra)
library(lme4)

rtdata <- read.csv("https://raw.githubusercontent.com/mkshaw/learn-mlms/master/data/hoffman2007.csv")
```


```{r}
#need to make salience a 01 type data structure; between 1 and 5 so maybe if painting salience is less than 2.5 and greater than equal to 2.5
rtdata <- rtdata %>%
          mutate(salience = ifelse(salience >= 2.5, 1,0))
```

```{r}
# Select numeric variables for the scatterplot matrix
numeric_vars <- rtdata %>%
                dplyr::select(age, salience, rt_sec, meaning, c_mean, c_sal)

# Create the scatterplot matrix
ggpairs(numeric_vars, 
        aes(color = as.factor(salience), alpha = 0.7)) +
        theme_minimal()
```

```{r}
#random intercepts model
RImodel <- lmer(rt_sec ~ 1 + (1|id), data = rtdata)

summary(RImodel)
```
```{r}
ggplot(data = data.frame(residuals = resid(RImodel)), aes(sample = residuals)) +
stat_qq() +
stat_qq_line() +
theme_minimal() +
labs(title = "QQ Plot of Residuals",
     x = "Theoretical Quantiles",
     y = "Sample Quantiles")
```

Inter-class correlation coefficient calculation
```{r}
var_hat <- 53.44
var_uhat <- 12.31
  
phat <- var_uhat/ (var_uhat+var_hat)
# the closer this number is to 1, the more evidence we have to use a multilevel model
print(phat)
```

Our fixed effect parameter was 7.4402, which represents the average reaction time across all individuals all ages.

Our inter-class correlation Coefficient Calculation showed that 18.72% of total variation in reaction time score is due to the difference in person.

Random Slopes and Intercept model

```{r}
RSImodel <- lmer(rt_sec ~ salience + (1 + salience|id), REML = TRUE, data = rtdata)
summary(RSImodel)
```

Based on the random slopes and intercepts model, we expect reaction time to decrease by 1.09984 on average for images with a salience greater than 2.5. The more salient a picture is means the more obvious the change will be so this indicates that as the image get clearer the reaction time decreases. 

```{r}
ggplot(data = data.frame(Fitted = fitted(RSImodel), Residuals = resid(RSImodel)), 
                         aes(x = Fitted, y = Residuals)) +
  
geom_point(alpha = 0.6, color = "#112446") +
geom_hline(yintercept = 0, color = "red") +
theme_minimal() +
labs(title = "Residuals vs. Fitted Values",
     x = "Fitted Values",
     y = "Residuals")
```

Model 3- REML = False
```{r}
RSImodel <- lmer(rt_sec ~ salience + oldage + oldage:salience + (1 + salience|id), REML = FALSE, data = rtdata)
summary(RSImodel)

AIC(RSImodel)
BIC(RSImodel)
```

```{r}
ggplot(data = data.frame(residuals = resid(RSImodel)), aes(sample = residuals)) +
stat_qq() +
stat_qq_line() +
theme_minimal() +
labs(title = "QQ Plot of Residuals",
     x = "Theoretical Quantiles",
     y = "Sample Quantiles")
```

